# Introduction

```{r setup, include = FALSE}
library(tidyverse)
library(modelsummary)

knitr::opts_chunk$set(echo = FALSE)
```


Choice models are an important part of transportation modeling and forecasting. 
One early application was by @train1978goods.

```
this is a block of code
```


We can also make lists

  - unnumbered lists
  - for things that don't need to be numbered
  

  
  
Or, we can have
  
  1. Ordered lists 
  2. For things that come in order.
  

We can [link](https://byu.edu) to webpages in a pretty simple way.


There are good ways to do equations, like the probability equation from an 
multinomial logit model.

\begin{equation}
  P_i = \frac{e^{V_i}}{\sum_{j \in J} e^{V_j}}
  (\#eq:mnl-probs)
\end{equation}

If you recall the choice probabilities in \@ref(eq:mnl-probs), you see.

Figure \@ref(fig:normplot) shows a normal cumulative probability distribution.

```{r normplot, fig.cap="Normal (Gaussian) cumulative probability distribution."}
tibble(
  x = seq(-3, 3, by = 0.01),
  y = pnorm(x)
) %>%
  ggplot(aes(x = x, y = y)) + geom_line()
```

We can also put in tables. I've really gotten to like the `modelsummary` package
and its functions. See an example in Table \@ref(tab:datasummary)


```{r datasummary}
data(trees)
models <- list()
models[['Bivariate']] <- lm(Girth ~ Height, data = trees)
models[['Multivariate']] <- lm(Girth ~ Height + Volume, data = trees)

# simple table
modelsummary(models, stars = TRUE, title = "Estimates of Tree Girth")
```
## Background

## Use of Disaggregate Discrete Choice Models

## Application Context in Current Course

## Urban and Intercity Travel Mode Choice Modeling

### Urban Travel Mode Choice Modeling

### Intercity Mode Choice Models

## Description of the Course

## Organization of the Course Structure