---
title: "Practice Exam"
author: "694R: Advanced Choice Modeling"
date: "4/6/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

> This is your practice exam. The final will have a similar format, but with a
different data set. All resources are available to you, including notes, the
course textbook, the internet, and your classmates. Your final exam will be the
same, *except* you must complete the final exam independently. As a result, I
encourage you to work as independently as possible, offering help but not
seeking it unless you are entirely stuck.

The `mlogit` library for R includes a dataset called `Car`,
which represents the car purchased by a person from a set of options.

```{r Car}
# load Car object into workspace
data(Car, package = "mlogit")
# show first ten rows of data
Car %>% as_tibble()
```

Each of the `r nrow(Car)` individuals in the data has six alternatives. A
complete data dictionary is available in the mlogit help files by calling
`?mlogit::Car`.  Transform this dataset as necessary to estimate choice models using the 
`mlogit()` function. 

Identify a preferred model to explain car choice. Consider the following
in your identification process:

  - Alternative representations of model parameters (e.g., log transforms and divisions).
  - Statistical significance and behavioral intuitiveness of the model parameters.
  - Comparative relationships between model parameters (e.g., value of time estimate).
  - Statistical goodness of fit tests between candidate models.
  - Data segmentation
  - Nesting structures

Construct your analysis in an Rmarkdown file, paired with appropriate textual
discussion. Expose your model code, but print your model results and statistical 
tests in publication-quality HTML tables. Submit your html file to the assignment
on Learning Suite before the deadline, after confirming that you can view the
file independently in a web browser.

